Model Finding for Recursive Functions in SMT\thanks{ This research is partially supported by the Inria technological development action ``Contre-exemples Utilisables par Isabelle et Coq'' (CUIC). }

### 2015-05-06 tex/main.tex
more

\newcommand\cvcd{CVC4d\xspace}
\newcommand\cvcf{CVC4f\xspace}
\newcommand\cvci{CVC4i\xspace}
\newcommand\ziiib{\ziii\xspace}

@@

Recent techniques~\cite{GeDeM-CAV-09, ReyEtAl-1-RR-13,reynolds-et-al-2013}
have focused on finding models for quantified formulas in SMT.
Ge and de Moura \cite{GeDeM-CAV-09} presented a complete instantiation-based
procedure for formulas in the \emph{essentially uninterpreted} fragment.
This fragment is limited to universally quantified formulas where all
occurrences of variables occur as direct subterms of uninterpreted
functions. For example, the formula $\forall x%\typ{\sortint} --- the sort is not important, right?
.\;\, \con{f}( x )
\teq \con{g}( x ) + 5$ is in the essentially uninterpreted fragment.
% since all
% occurrences of $x$ occur beneath uninterpreted functions $\con{f}$ and
% $\con{g}$.
Other syntactic criteria are identified in this paper that extend
this fragment slightly, including cases when variables occur as arguments of
arithmetic predicates. The techniques in~Reynolds et al.\
\cite{ReyEtAl-1-RR-13,reynolds-et-al-2013} can be used for finding finite
models for quantified
formulas over uninterpreted types as well as types having a fixed finite
interpretation, such as fixed-width bitvectors and enumerated datatypes.
Finite model finding techniques can
find a model for the formula $\forall x,\, y : \tau.\;\, x \teq
y \vee \neg \con{f}( x ) \teq \con{f}( y )$ where the uninterpreted type $\tau$ is
interpreted as a finite set and $\con{f}$ as injective finite map.

@@

A \emph{signature} $\Sigma$ consists of
a set $\ssorts{\Sigma}$ of type symbols (or sorts) and
a set $\sfuns{\Sigma}$ of function symbols.
We assume that signatures always include a Boolean type $\Bool$ and constants
$\ltrue$ (truth) and $\lfalse$ (falsity) of that type,
an infix equality predicate ${\teq} : \tau \times \tau \to \Bool$
for each $\tau \in \ssorts{\Sigma}$,
standard boolean connectives ($\neg$, $\wedge$, $\vee$, etc.),
and a function $\lite : \Bool \times \tau \times \tau \rightarrow \tau$ for each $\tau \in \ssorts{\Sigma}$.
We write $t\typ{\tau}$ to denote a term $t$ of type~$\tau$ and
$\terms( t )$ to denote the set of subterms in $t$.
%We write $\functerms^\Sigma( t )$ to denote the set of $f$-applications in $t$ such that $f \in \sfundefs{\Sigma}$.

@@

We gathered a set of 245 benchmarks from two sources, which we will refer to as
{\bench isa} and {\bench leon}. The first ({\bench isa}) consists of the 79
benchmarks from the Isaplanner benchmark
suite~\cite{DBLP:conf/itp/JohanssonDB10} that do not contain higher-order
functions. The benchmarks have been used recently as challenge problems for a
variety of inductive theorem provers. These benchmarks heavily involve
recursive functions, and are limited to the combined theory of uninterpreted
functions and inductive datatypes. The second ({\bench leon}) consists of 166
benchmarks from the Leon repository,\footnote{%Available at
\url{https://github.com/epfl-lara/leon}} which were constructed from
verification conditions from simple Scala programs. These benchmarks also
heavily involve recursive functions over inductive datatypes, and also cover a
wide variety of theories, including bitvectors, arrays, and linear and
non-linear arithmetic. Each of the 245 benchmarks are of the form described in
formula~(\ref{eq:wdf}) from Section~\ref{sec:prelim}.
% in particular they each
%are a conjoined list of function definitions followed by a (negated)
%conjecture.
A majority of these benchmarks are unsatisfiable, although a
handful of the benchmarks in the {\bench leon} set are satisfiable.

@@

The results are shown in Figures \ref{fig:sat} and \ref{fig:unsat}.
The results are divided into benchmarks triggering ``unsatisfiable'' and ``satisfiable''
responses and further into benchmarks before and after the translation $\conv$.
\jb{rephrase}
First, we note that in no cases did a configuration answer differently across the two encodings.
In other words, there were no cases in which \ziii or \cvc answered e.g.\ ``unsatisfiable'' on a benchmark $\varphi$
and ``satisfiable'' on its corresponding benchmark $\conv( \varphi )$, or vice versa.
This provides confirming evidence that the translation $\conv$ is \emph{refutation-sound},
as we did not find a case where $\conv( \varphi )$ was unsatisfiable when $\varphi$ was satisfiable.
It also provides confirming evidence that the translation is \emph{model-sound} for this set of benchmarks,
as we did not find a case where $\conv( \varphi )$ was satisfiable when $\varphi$ was unsatisfiable.
This is to be expected, since the functions in our benchmark sets are expected to be admissible (and in fact, terminating).

Examining the first encoding in isolation (the first nine columns),
the number of satisfiable responses from all configurations is quite low.
In total, \ziiib manages to find only 11 benchmarks to be satisfiable,
whereas each configuration of \cvc only finds 6.
This confirms the shortcomings of existing SMT techniques for finding counterexamples for benchmarks containing recursive functions.
In terms of number of unsatisfiable responses,
the configuration \cvci is the clear winner, finding 318 total benchmarks to be unsatisfiable.

Now, examining the second encoding (the latter nine columns),
we see that using the translation $\conv$ drastically improves the ability of configurations for answering satisfiable.
In total, \cvcf finds 331 of the 885 benchmarks to be satisfiable, including 9 benchmarks in the original {\bench leon} benchmark set.
The performance of \ziiib for counterexamples also improves drastically, as it solves 112 satisfiable,
including 10 that were unsolved by \cvcf.
Moreover, the translation $\conv$ helps \ziii for unsatisfiable responses as well,
as we see \ziii solves a total of 209 whereas it solves only 187 before the translation.
This number is 2 fewer than \cvcd, which solves 211 unsatisfiable.
Thus, in most cases the translation $\conv$ does not significantly degrade performance for unsatisfiable benchmarks,
and in some cases may actually aid the solver for determining unsatisfiability.%
\footnote{The exception here is \cvci, which makes use of type
information associated with quantified formulas for inductive arguments.}

We conclude from this evaluation that the translation $\conv$ enables SMT solvers to find counterexamples
for conjectures involving recursive functions whose definitions are admissible.
Both \ziiib and \cvcf were able to find a large number of counterexamples after the translation,
with the latter configuration finding significantly more instances (331 to 112).

>>>>

\newcommand\cvcd{CVC4d\xspace}
\newcommand\cvcf{CVC4f\xspace}
\newcommand\cvci{CVC4i\xspace}
\newcommand\ziiib{\ziii}

@@

Recent techniques~\cite{GeDeM-CAV-09, ReyEtAl-1-RR-13,reynolds-et-al-2013}
have focused on finding models for quantified formulas in SMT.
Ge and de Moura \cite{GeDeM-CAV-09} presented a complete instantiation-based
procedure for formulas in the \emph{essentially uninterpreted} fragment.
This fragment is limited to universally quantified formulas where all
occurrences of variables occur as direct subterms of uninterpreted
functions. For example, the formula $\forall x%\typ{\sortint} --- the sort is not important, right?
.\;\, \con{f}( x )
\teq \con{g}( x ) + 5$ is in the essentially uninterpreted fragment.
% since all
% occurrences of $x$ occur beneath uninterpreted functions $\con{f}$ and
% $\con{g}$.
Other syntactic criteria are identified in this paper that extend
this fragment slightly, including cases when variables occur as arguments of
arithmetic predicates. The techniques in~Reynolds et al.\
\cite{ReyEtAl-1-RR-13,reynolds-et-al-2013} can be used for finding finite
models for quantified
formulas over uninterpreted types as well as types having a fixed finite
interpretation, such as fixed-width bit vectors and enumerated datatypes.
Finite model finding techniques can
find a model for the formula $\forall x,\, y : \tau.\;\, x \teq
y \vee \neg \con{f}( x ) \teq \con{f}( y )$ where the uninterpreted type $\tau$ is
interpreted as a finite set and $\con{f}$ as injective finite map.

@@

A \emph{signature} $\Sigma$ consists of
a set $\ssorts{\Sigma}$ of type symbols (or sorts) and
a set $\sfuns{\Sigma}$ of function symbols.
We assume that signatures always include a Boolean type $\Bool$ and constants
$\ltrue$ (truth) and $\lfalse$ (falsity) of that type,
an infix equality predicate ${\teq} : \tau \times \tau \to \Bool$
for each $\tau \in \ssorts{\Sigma}$,
standard boolean connectives ($\neg$, $\wedge$, $\vee$, etc.),
and a function $\lite : \Bool \times \tau \times \tau \rightarrow \tau$ for each $\tau \in \ssorts{\Sigma}$.
We write $t\typ{\tau}$ to denote a metavariable ranging over terms of
type~$\tau$ and $\terms( t )$ to denote the set of subterms in $t$.
%We write $\functerms^\Sigma( t )$ to denote the set of $f$-applications in $t$ such that $f \in \sfundefs{\Sigma}$.

@@

We gathered a set of 245 benchmarks from two sources, which we will refer to as
{\bench isa} and {\bench leon}. The first ({\bench isa}) consists of the 79
benchmarks from the Isaplanner benchmark
suite~\cite{DBLP:conf/itp/JohanssonDB10} that do not contain higher-order
functions. The benchmarks have been used recently as challenge problems for a
variety of inductive theorem provers. These benchmarks heavily involve
recursive functions, and are limited to the combined theory of uninterpreted
functions and inductive datatypes. The second ({\bench leon}) consists of 166
benchmarks from the Leon repository,\footnote{%Available at
\url{https://github.com/epfl-lara/leon}} which were constructed from
verification conditions from simple Scala programs. These benchmarks also
heavily involve recursive functions over inductive datatypes, and also cover a
wide variety of theories, including bit vectors, arrays, and linear and
nonlinear arithmetic. Each benchmark is in definitional form with respect to a
set of terminating functions.
%
% in particular they each
%are a conjoined list of function definitions followed by a (negated)
%conjecture.
A majority of these benchmarks are unsatisfiable, although a
handful of the benchmarks in the {\bench leon} set are satisfiable.

@@

The results are shown in Figures \ref{fig:sat} and \ref{fig:unsat}.
The results are divided into benchmarks triggering ``unsatisfiable'' and ``satisfiable''
responses and further into benchmarks before and after the translation $\conv$.
\jb{rephrase}
%
The raw evaluation data reveals no cases in which a solver answered
``unsatisfiable'' on a benchmark $\varphi$ and ``satisfiable'' on its
corresponding benchmark $\conv( \varphi )$, or vice versa, thereby
corroborating Theorem~\ref{thm:equi}.

Figure~\ref{fig:sat} shows that for untranslated benchmarks (the ``$\varphi$''
columns), the number of satisfiable responses is very low, across all
configurations. This confirms the shortcomings of existing SMT techniques for
finding counterexamples for benchmarks containing recursive functions.
%
The translation $\conv$ (the ``$\conv(\varphi)$'' columns) has a major
impact. \cvcf finds 331 of the 885 benchmarks to be satisfiable,
including 9~benchmarks in the unmutated {\bench leon} benchmark set. The
performance of \ziiib for counterexamples also improves dramatically, as it
solves 112 satisfiable problems, including 10 that were unsolved by \cvcf.
%
We conclude that the translation $\conv$ enables SMT
solvers to find counterexamples for conjectures involving recursive functions
whose definitions are admissible.
%Both \ziiib and \cvcf were able to find a
%large number of counterexamples after the translation, with the latter
%configuration finding significantly more instances (331 to 112).

In terms of number of unsatisfiable responses, the configuration \cvci is the
clear winner, finding 318 total benchmarks to be unsatisfiable. Moreover, the
translation $\conv$ helps \ziii for ``unsatisfiable'' responses as well, as we
see \ziii solves a total of 209 whereas it solves only 187 before the
translation.
%
This number is 2 fewer than \cvcd, which solves 211 unsatisfiable. Thus, in
most cases the translation $\conv$ does not significantly degrade performance
for unsatisfiable benchmarks, and in some cases may actually aid the solver for
determining unsatisfiability. The exception here is \cvci, whose inductive
arguments are guided by the types of quantified variables.
